Normalization by Evaluation for Sized Dependent Types

### 2017-02-20 nbe/icfp17.tex
[ ICFP 17 ] Overview over NbE.

@@

@@

@@

\section{Completeness of NbE}
\label{sec:compl}

>>>>

\usepackage{mathtools} %mathrlap
\usepackage[all]{xy}

@@

\begin{figure}
$$
\xymatrix@R=12ex@C=12ex{
\mathrlap{\mbox{Semantics ($\beta$)}} &
  &                f \in \D   \ar@[red][d]_{\down A}
\\
\mathrlap{\mbox{Semantics ($\beta\eta$)}} &
  &                d \in \DNf \ar@[red][d]_{\tR_k{}}
  &                n \in \DNe \ar[d]_{\tRne_k} \ar@[blue][ul]_{\up A}
  & \ar@[blue][l]_{\var{}} j \in \Level \ar@{<->}[d]_{k\dotminus(1+\_)}
\\
\mathrlap{\mbox{Syntax}} &
t \in \Exp \ar@[red]@(u,dl)[uur]^{\_\color{blue}\eta} %\ar@/^/[uur]^{\den\_}
  & \ar[l]_{\supseteq}  v \in \Nf
  & \ar[l]_{\supseteq}  m \in \Ne
  & \ar[l]_{\ind{}}     i \in \Index
  &
}
$$
\caption{Type-assignment NbE in locally nameless
    style.\label{fig:typedchart}}
\end{figure}

Figure~\ref{fig:typedchart}, adapted from \citet{abel:habil} shows the syntactic categories and main operations involved in NbE in what we call \emph{locally nameless style}.  The red path $\Exp \to \D \to \Dnf \to \Nf$ decomposes $\beta\eta$-normalization into three steps.
\begin{enumerate}
\item
First, we close the term $t$ with an environment $\eta$ that maps the free de Bruijn indices of $t$ to reflected de Bruijn levels.  Reflection of de Bruijn levels follows the blue path $\Level \to \Dne \to \D$: Levels embed via constructor $\var{}$ into semantic neutrals $\DNe$ which are labelled with their type $A \in \D$ to become and element $\up A \lev j \in \D$.

\item
Then, we label value $t \eta \in \D$ with its type $A$ to obtain $\down A t \eta \in \Dnf$.
\item
Finally, \emph{read back} %% $\tR_k$
$\tR_k \down{A} t\eta$ produces a long normal form $v \in \Nf$, converting de Bruijn levels back to indices.
Herein, $k$ should be the length of the context the original term $t$ lived in.
If this is the case, each de Bruijn level encountered during read back is below $k$ and can be safely converted to a de Bruijn index.
\end{enumerate}

@@

Note that $T \in \SET[\ell]$ and $T \shape S$ do not imply $S \in \SET[\ell]$.
Type shapes are not well-defined types in general.
For instance, assume a term $F : \Nat 0 \to \Set[0]$ which diverges if applied to a successor term.
Then $T := \funT x {\Nat 0} {F\,x}$ is a welldefined type; we have $T \in \SET[0]$.
Now consider $S := \funT x {\Nat \infty} {F\,x}$.
We have $T \shape S$, but not $S$ is not well-defined; $S \not\in \SET[0]$.
% $\funT x {\Nat 0} {F\,x} \shape , but the latter one is not a well-defined type.

@@

\subsection{Completeness of NbE}
\label{sec:compl}

\section{Soundness  of NbE}
\label{sec:compl}

In this section, we show that NbE is sound for judgemental equality, i.e., that same normal form implies definitional equality.  The proof follows previous work on NbE \cite{abelCoquandDybjer:lics07,abel:habil}.

\[
  \fbox{$\LRSize \Gamma a b$} \defas
  \forall \sigma : \Gamma' \leq \Gamma.\
  \exists a'.\
  \RSize {\Gamma'} b a' \mand
  \Gamma' \der a\sigma = a' : \Size
\]

\fbox{$\LR \Gamma t {T'} f {A'}$} defined by induction on ${A'} \in \SET[\ell]$.
\begin{caselist}

\nextcase $A' \evalsto \piT A B$.  Then $\LR \Gamma t {T'} f {A'}$ if all of the following:
\[
\begin{array}{l}
\Gamma \der T' = \piT U T \mforsome U, T\\
\LRT \Gamma U A \\
\forall \sigma : \Gamma' \leq \Gamma.\ \
\LR {\Gamma'} u {U\sigma} a A \implies \LR {\Gamma'} {t\sigma\,u} {T(\sigma,u)} {f\,a} {B[a]}
\end{array}
\]

\nextcase $A' \evalsto \forallT B$.  Then $\LR \Gamma t {T'} f {A'}$ if all of the following:
\[
\begin{array}{l}
\Gamma \der T' = \forallT T \mforsome T\\
\forall \sigma : \Gamma' \leq \Gamma.\ \
\LRSize {\Gamma'} u a \implies \LR {\Gamma'} {t\sigma\,u} {T(\sigma,u)} {f\,a} {B[a]}
\end{array}
\]

\end{caselist}
